---
title: "Documentation"
author: Ben Wieland
date: Sys.Date()
output: md_document
---

# fRisbee: For college ultimate frisbee

fRisbee is an R package designed to make interfacting with college frisbee team rankings straightforward & easy. 

Install the package using the code below.

```{r results = F, message = F, warning = F}
devtools::install_github("https://github.com/bbwieland/fRisbee",upgrade = "ask",force = F)
library(fRisbee)
```

We can load the most up-to-date college rankings from www.frisbee-rankings.com 

``` {r warning = F}
head(fRisbee::GetFrisbeeRankings())
```

Using this data, lots of interesting analysis can be performed — for example, this scatterplot of D-I strength of schedule vs. team rating: 

(Note: you'll need to install the ggrepel package to replicate this!)

```{r}
library(ggrepel)

data = fRisbee::GetFrisbeeRankings(DivisionIOnly = T) %>%
  filter(Rating >= 0)

ggplot(data,aes(x = SoS,y = Rating,label = Team)) +
  geom_point() +
  geom_smooth(method = "lm", se = F,linetype = "dashed") +
  geom_text_repel(max.overlaps = 8) +
  annotate("label",x = 750, y = 2000, label = "Overperform vs. schedule",size = 4) +
  annotate("label",x = 1500, y = 500, label = "Underperform vs. schedule",size = 4) +
  theme_classic() +
  labs(title = "How does each team perform, relative to its strength of schedule?",
       x = "Strength of schedule (average opp. rating)",
       y = "Team rating")
```

Using fRisbee, it is also possible to scrape all of a team's games played in the season. 

```{r}
## scraping all Virginia games for the season

virginia = fRisbee::GetTeamResults("Virginia")

# Who were Virginia's ten toughest opponents this season, and how did they perform against them? 

virginia %>% 
  dplyr::arrange(OppRk) %>% 
  head(.,10)

# How did Virginia's performance vary by opponent? 

ggplot(virginia,aes(x = OpponentRating, y = PtDiff)) +
  geom_point() + 
  scale_x_continuous(limits = c(500,2500)) +
  geom_smooth(method = "lm",se = F) +
  geom_hline(yintercept = 0,color = "red", linetype = "dashed") +
  theme_classic() +
  labs(title = "Virginia team performance by opponent strength",
       subtitle = "As expected, point differential decreases against stronger opponents.",
       x = "Opponent rating at time of game",
       y = "Point differential")

```


The USA Ultimate Frisbee website specifices an algorithm for determining score for each game, developed such that the minimum gain in score for any team is 125 points (after a one-goal win) and the maximum gain in score for any team is 600 points (if they more than double their opponent's score). The GameScoreCalculator function computes the game score for any given winner and loser scores. 

```{r}
# A close one-goal victory results in the minimum rating change of 125 points.
fRisbee::GameScoreCalculator(13,12)

# A blowout results in the maximum rating change of 600 points.
fRisbee::GameScoreCalculator(13,3)

# Let's look at how each possible outcome affects game score. 
# The x-axis displays potential opponent scores in a game won with 13 points.

seq = 12:1

scores = seq %>%
  map(~GameScoreCalculator(winner_score = 13, loser_score = .x)) %>%
  unlist()

plot_data = data.frame(opp_points = seq, score = scores)

ggplot(plot_data,aes(opp_points,score)) +
  scale_x_continuous(limits = c(0,12),breaks = seq(1,12,1)) +
  geom_line() +
  theme_classic() +
  labs(x = "Opponent Points Scored",
       y = "Game Score",
       title = "Game score performance of a team that wins with 13 points")

```

The ratings change produced by a specific game is a function of game score AND the opponent's rating coming into a game. Basically, if your team's game score exceeds the difference between your team's rating and your opponent's rating, your rating will INCREASE as a result; otherwise, it'll DECREASE! 

Yes — this means that wins don't always result in ratings increases! You need to beat a team by **more** than your "expected" win to see a ratings increase; the inverse is also true: a weaker team that holds its own against a strong team may see a ratings increase. 

Use the RatingAdjustedGameScoreCalculator to determine how, given each team's initial rating and each team's game performance, their final rating will be affected by the outcome.

```{r}
# In an evenly matched game, a close victory results in a ratings increase.
fRisbee::RatingAdjustedGameScoreCalculator(winner_rating = 1500,loser_rating = 1500,winner_score = 13, loser_score = 12)
```

``` {r}
# However, a highly-favored team winning a close game actually results in a ratings increase for the LOSING team.
fRisbee::RatingAdjustedGameScoreCalculator(winner_rating = 1500,loser_rating = 1000,winner_score = 13, loser_score = 12)
```

We can combine two functions, GetFrisbeeRankings and RatingAdjustedGameScoreCalculator, to create a useful composite function: RatingAdjustedGameScoreCalculatorTeam.

This works similary to RatingAdjustedGameScoreCalculator, but it takes team names as inputs instead of ratings. The function automatically pulls the most recent ratings for those teams, and calculates the game score change accordingly.

```{r}
fRisbee::RatingAdjustedGameScoreCalculatorTeam(winner_team = "Virginia",loser_team = "Virginia Tech",winner_score = 13, loser_score = 8)
```

All data in this vignette as of May 7, 2022. 
Data sourced from frisbee-rankings.com — thanks to Cody Mills for developing the rankings site!
